TG2 Publications

### 2025-04-24 papers.qmd
Update papers.qmd content with additional publications

@@

Willi Sauerbrei,  Aris Perperoglou,  Matthias Schmid et al  for TG2 of the STRATOS initiative

[Link](https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)

@@

Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid on behalf of TG2 of the STRATOS initiative

[Link](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)

<details>
  <summary>Abstract</summary>
### Background

@@

[Link](https://stratos-initiative.org/sites/default/files/STRATOS-TG2-BB-variable-selection.pdf)

@@

Paul Bach, Christine Wallisch, Nadja Klein, Lorena Hafermann, Willi Sauerbrei, Ewout W. Steyerberg, Georg Heinze, Geraldine Rauch , for topic group 2 of the STRATOS initiative

[Link](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)

>>>>

## Evaluating variable selection methods for multivariable regression models: A simulation study protocol

Ullmann, T., Heinze, G., Hafermann, L., Schilhart-Wallisch, C., Dunkler, D., for TG2 of the STRATOS initiative, 2024. Evaluating variable selection methods for multivariable regression models: A simulation study protocol. PLoS ONE 19, e0308543. [https://doi.org/10.1371/journal.pone.0308543](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0308543)

### Abstract

Researchers often perform data-driven variable selection when modeling the associations between an outcome and multiple independent variables in regression analysis. Variable selection may improve the interpretability, parsimony and/or predictive accuracy of a model. Yet variable selection can also have negative consequences, such as false exclusion of important variables or inclusion of noise variables, biased estimation of regression coefficients, underestimated standard errors and invalid confidence intervals, as well as model instability. While the potential advantages and disadvantages of variable selection have been discussed in the literature for decades, few large-scale simulation studies have neutrally compared data-driven variable selection methods with respect to their consequences for the resulting models. We present the protocol for a simulation study that will evaluate different variable selection methods: forward selection, stepwise forward selection, backward elimination, augmented backward elimination, univariable selection, univariable selection followed by backward elimination, and penalized likelihood approaches (Lasso, relaxed Lasso, adaptive Lasso). These methods will be compared with respect to false inclusion and/or exclusion of variables, consequences on bias and variance of the estimated regression coefficients, the validity of the confidence intervals for the coefficients, the accuracy of the estimated variable importance ranking, and the predictive performance of the selected models. We consider both linear and logistic regression in a low-dimensional setting (20 independent variables with 10 true predictors and 10 noise variables). The simulation will be based on real-world data from the National Health and Nutrition Examination Survey (NHANES). Publishing this study protocol ahead of performing the simulation increases transparency and allows integrating the perspective of other experts into the study design.

</details>

## Regression without regrets –initial data analysis is a prerequisite for multivariable regression

Heinze, G., Baillie, M., Lusa, L., Sauerbrei, W., Schmidt, C.O., Harrell, F.E., Huebner, M., on behalf of TG2 and TG3 of the STRATOS initiative, 2024. Regression without regrets –initial data analysis is a prerequisite for multivariable regression. BMC Med Res Methodol 24, 178. [https://doi.org/10.1186/s12874-024-02294-3](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-024-02294-3)

### Abstract

Statistical regression models are used for predicting outcomes based on the values of some predictor variables or for describing the association of an outcome with predictors. With a data set at hand, a regression model can be easily fit with standard software packages. This bears the risk that data analysts may rush to perform sophisticated analyses without sufficient knowledge of basic properties, associations in and errors of their data, leading to wrong interpretation and presentation of the modeling results that lacks clarity. Ignorance about special features of the data such as redundancies or particular distributions may even invalidate the chosen analysis strategy. Initial data analysis (IDA) is prerequisite to regression analyses as it provides knowledge about the data needed to confirm the appropriateness of or to refine a chosen model building strategy, to interpret the modeling results correctly, and to guide the presentation of modeling results. In order to facilitate reproducibility, IDA needs to be preplanned, an IDA plan should be included in the general statistical analysis plan of a research project, and results should be well documented. Biased statistical inference of the final regression model can be minimized if IDA abstains from evaluating associations of outcome and predictors, a key principle of IDA. We give advice on which aspects to consider in an IDA plan for data screening in the context of regression modeling to supplement the statistical analysis plan. We illustrate this IDA plan for data screening in an example of a typical diagnostic modeling project and give recommendations for data visualizations.

</details>

## Review of guidance papers on regression modeling in statistical series of medical journals

Wallisch, C., Bach, P., Hafermann, L., Klein, N., Sauerbrei, W., Steyerberg, E.W., Heinze, G., Rauch, G., on behalf of topic group 2 of the STRATOS initiative, 2022. Review of guidance papers on regression modeling in statistical series of medical journals. PLoS ONE 17, e0262918. [https://doi.org/10.1371/journal.pone.0262918](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0262918)

### Abstract

Although regression models play a central role in the analysis of medical research projects, there still exist many misconceptions on various aspects of modeling leading to faulty analyses. Indeed, the rapidly developing statistical methodology and its recent advances in regression modeling do not seem to be adequately reflected in many medical publications. This problem of knowledge transfer from statistical research to application was identified by some medical journals, which have published series of statistical tutorials and (shorter) papers mainly addressing medical researchers. The aim of this review was to assess the current level of knowledge with regard to regression modeling contained in such statistical papers. We searched for target series by a request to international statistical experts. We identified 23 series including 57 topic-relevant articles. Within each article, two independent raters analyzed the content by investigating 44 predefined aspects on regression modeling. We assessed to what extent the aspects were explained and if examples, software advices, and recommendations for or against specific methods were given. Most series (21/23) included at least one article on multivariable regression. Logistic regression was the most frequently described regression type (19/23), followed by linear regression (18/23), Cox regression and survival models (12/23) and Poisson regression (3/23). Most general aspects on regression modeling, e.g. model assumptions, reporting and interpretation of regression results, were covered. We did not find many misconceptions or misleading recommendations, but we identified relevant gaps, in particular with respect to addressing nonlinear effects of continuous predictors, model specification and variable selection. Specific recommendations on software were rarely given. Statistical guidance should be developed for nonlinear effects, model specification and variable selection to better support medical researchers who perform or interpret regression analyses.

</details>

Bach, P., Wallisch, C., Klein, N., Hafermann, L., Sauerbrei, W., Steyerberg, E.W., Heinze, G., Rauch, G., for topic group 2 of the STRATOS initiative, 2020. Systematic review of education and practical guidance on regression modeling for medical researchers who lack a strong statistical background: Study protocol. PLoS ONE 15, e0241427. [https://doi.org/10.1371/journal.pone.0241427](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)

@@

Sauerbrei, W., Perperoglou, A., Schmid, M., Abrahamowicz, M., Becher, H., Binder, H., Dunkler, D., Harrell, F.E., Royston, P., Heinze, G., for TG2 of the STRATOS initiative,  2020. State of the art in selection of variables and functional forms in multivariable analysis—outstanding issues. Diagn Progn Res 4, 3, s41512-020-00074–3. [https://doi.org/10.1186/s41512-020-00074-3](https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)

@@

Perperoglou, A., Sauerbrei, W., Abrahamowicz, M., Schmid, M., 2019. A review of spline function procedures in R. BMC Med Res Methodol 19, 46. [https://doi.org/10.1186/s12874-019-0666-3](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
<details>
  <summary>Abstract</summary>
### Background

@@

[Link](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG2-BB-variable-selection.pdf)

@@
